Finally Tagless, Partially Evaluated

### 2007-06-11 tagless/tagless-final.tex
just some tricks to reduce space

\begin{figure}
%
(a)\vspace{-10pt}
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool | L of exp | A of exp * exp
\end{code}
%
(b)\vspace{-10pt}
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}
%
(c)\vspace{-10pt}
\begin{code}
let rec lookup (x::env) = function
| VZ   -> x
| VS v -> lookup env v

let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}
%
(d)\vspace{-10pt}
\begin{code}
type u = UB of bool | UA of (u -> u)
\end{code}
%
(e)\vspace{-10pt}
\begin{code}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code}
%
(f)\vspace{-10pt}
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
%
(g)\vspace{-10pt}
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
%
(h)\vspace{-10pt}
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
%
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
It is straightforward to create an algebraic data type, say in OCaml%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
in Fig.~\ref{fig:tag-problem}(b)%
% fig:tag-problem part (b) was here

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(a)%
.
% fig:tag-final-proposal part (a) was here
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
Fig.~\ref{fig:tag-final-proposal}(b)%
.
% fig:tag-final-proposal part (b) was here
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(c)%
.
% fig:tag-final-proposal part (c) was here
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(d)%
.
% fig:tag-final-proposal part (d) was here
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.
\begin{figure}
%
(a)\vspace{-10pt}
\begin{code}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
%
(b)\vspace{-10pt}
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
%
(c)\vspace{-10pt}
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
%
(d)\vspace{-10pt}
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
%
\caption{OCaml code illustrating our final proposal.}
\label{fig:tag-final-proposal}
\end{figure}
%
In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner 
type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a)\vspace{-10pt}
\begin{code}
class Symantics repr where
  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

@@

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{code}
%
(b)\vspace{-10pt}
\begin{code}
testpowfix () = lam (\x -> fix (\self -> lam (\n ->
                 if_ (leq n (int 0)) (int 1)
                     (mul x (app self (add n (int (-1))))))))
\end{code}
%
(c)\vspace{-10pt}
\begin{code}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}
%
\caption{Symantics in Haskell.}
\label{fig:symantics-haskell}
\end{figure}
% fig:symantics-haskell part (a) was here
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For another example, the classical $\mathit{power}$ function is
%aplas-specific
in Fig.~\ref{fig:symantics-haskell}(b)
% fig:symantics-haskell part (b) was here
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
in Fig.~\ref{fig:symantics-haskell}(c).
% fig:symantics-haskell part (c) was here
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr-> (unit-> ('c, 'da) repr)-> (unit-> ('c, 'da) repr)-> ('c, 'da) repr
end
\end{code}
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
                         if_ (leq n (int 0)) (fun () -> int 1)
                             (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(a)
.
% fig:interpreter-R part (a) was here
%
\begin{figure}
%
(a)\vspace{-10pt}
\begin{code}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

@@

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
%
(b)\vspace{-10pt}
\begin{code}
module EXR = EX(R)
\end{code}
%
(c)\vspace{-10pt}
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
%
\caption{The tagless interpreter \texttt{R}.}
\label{fig:interpreter-R}
\end{figure}
%
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(b)
.
% fig:interpreter-R part (b) was here
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
%aplas-specific
 Fig.~\ref{fig:interpreter-R}(c)
.
% fig:interpreter-R part (b) was here
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(a)
.
% fig:interpreter-C part (a) was here
%
\begin{figure}[t]
%
(a)\vspace{-10pt}
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

@@

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{code}
%
(b)\vspace{-10pt}
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
%
(c)\vspace{-10pt}
\begin{code}
let module E = EX(C) in E.testpowfix7
\end{code}
%
(d)\vspace{-10pt}
\begin{code}
.<fun x_1 -> (fun x_2 ->
  let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code}
%
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d)
% fig:interpreter-C part (d) was here

>>>>

\begin{figure}
%
(a) \begin{code2}
type var = VZ | VS of var
type exp = V of var | B of bool | L of exp | A of exp * exp
\end{code2}

(b) \begin{code2}
let test1 = A (L (V VZ), B true)
\end{code2}

(c) \begin{code2}
let rec lookup (x::env) = function
| VZ   -> x
| VS v -> lookup env v

let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code2}

(d) \begin{code2}
type u = UB of bool | UA of (u -> u)
\end{code2}

(e) \begin{code2}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code2}

(f) \begin{code2}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code2}

(g) \begin{code2}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code2}

(h) \begin{code2}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code2}
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
It is straightforward to create an algebraic data type, say in OCaml%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
in Fig.~\ref{fig:tag-problem}(b)%
% fig:tag-problem part (b) was here

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(a)%
.
% fig:tag-final-proposal part (a) was here
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
Fig.~\ref{fig:tag-final-proposal}(b)%
.
% fig:tag-final-proposal part (b) was here
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(c)%
.
% fig:tag-final-proposal part (c) was here
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error
%aplas-specific
, Fig.~\ref{fig:tag-final-proposal}(d)%
.
% fig:tag-final-proposal part (d) was here
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.
\begin{figure}
%
(a) \begin{code2}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code2}

(b) \begin{code2}
let testf1 = app (lam varZ) (b true)
\end{code2}

(c) \begin{code2}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code2}

(d) \begin{code2}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code2}
\caption{OCaml code illustrating our final proposal.}
\label{fig:tag-final-proposal}
\end{figure}
%
In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner 
type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

@@

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

@@

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{code2}

(b) \begin{code2}
testpowfix () = lam (\x -> fix (\self -> lam (\n ->
                 if_ (leq n (int 0)) (int 1)
                     (mul x (app self (add n (int (-1))))))))
\end{code2}

(c) \begin{code2}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code2}
\caption{Symantics in Haskell.}
\label{fig:symantics-haskell}
\end{figure}
% fig:symantics-haskell part (a) was here
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For another example, the classical $\mathit{power}$ function is
%aplas-specific
in Fig.~\ref{fig:symantics-haskell}(b)
% fig:symantics-haskell part (b) was here
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
in Fig.~\ref{fig:symantics-haskell}(c).
% fig:symantics-haskell part (c) was here
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr-> (unit-> ('c, 'da) repr)-> (unit-> ('c, 'da) repr)-> ('c, 'da) repr
end

module EX(S: Symantics) = struct
  open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =  lam (fun x -> fix (fun self -> lam (fun n ->
                         if_ (leq n (int 0)) (fun () -> int 1)
                             (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 =  lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

@@

The first interpreter evaluates an object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(a)
.
% fig:interpreter-R part (a) was here
%
\begin{figure}
(a) \begin{code2}
module R = struct
  type ('c,'dv) repr = 'dv (* no wrappers *)

@@

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code2}

(b) \begin{code2}
module EXR = EX(R)
\end{code2}

(c) \begin{code2}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code2}
\caption{The tagless interpreter \texttt{R}.}
\label{fig:interpreter-R}
\end{figure}
%
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with |R|%
%aplas-specific
, Fig.~\ref{fig:interpreter-R}(b)
.
% fig:interpreter-R part (b) was here
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
%aplas-specific
 Fig.~\ref{fig:interpreter-R}(c)
.
% fig:interpreter-R part (b) was here
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\begin{proposition}
If an object term~$e$ encoded in the metalanguage has type~$t$,
then evaluating~$e$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\begin{proposition}
  If an implementation of |Symantics| never gets stuck, then
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
\end{proposition}
These propositions follow immediately from the soundness of the
metalanguage's type system.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C| below%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(a)
.
% fig:interpreter-C part (a) was here
%
\begin{figure}[t]
% this code does not have the 'sv parameter. It shows up later.
(a) \begin{code2}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code

@@

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f =  .<let rec self n = .~(f .<self>.) n in self>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee  = .<if .~eb then .~(et ()) else .~(ee ())>.
end
\end{code2}

(b) \begin{code2}
let module E = EX(C) in E.test1 ()
\end{code2}

(c) \begin{code2}
let module E = EX(C) in E.testpowfix7
\end{code2}

(d) \begin{code2}
.<fun x_1 -> (fun x_2 ->
  let rec self_3 = fun n_4 ->
   (fun x_5 -> if x_5 <= 0 then 1  else x_2 * (self_3 (x_5 + (-1))))
   n_4 in self_3) x_1 7>.
\end{code2}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
%
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1|%
%aplas-specific
, Fig.~\ref{fig:interpreter-C}(b),
% fig:interpreter-C part (b) was here
gives the code value |.<(fun x_6 -> x_6) true>.|
of inferred type |('c, bool) C.repr|.  Interpreting |testpowfix7|
with
%aplas-specific
Fig.~\ref{fig:interpreter-C}(c),
% fig:interpreter-C part (c) was here
gives a code value with many apparent $\beta$-redexes
%aplas-specific
Fig.~\ref{fig:interpreter-C}(d)
% fig:interpreter-C part (d) was here
